---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# rEDM
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/grand-total/rEDM)](https://cran.rstudio.com/web/packages/rEDM/index.html)

## Overview

The `rEDM` package is a collection of methods for Empirical Dynamic Modeling (EDM). EDM is based on the mathematical theory of reconstructing attractor manifolds from time series data, with applications to forecasting, causal inference, and more. It is based on research software previously developed for the Sugihara Lab (University of California San Diego, Scripps Institution of Oceanography).

## Empirical Dynamic Modeling (EDM)
---
This package implements an R wrapper of [EDM](http://deepeco.ucsd.edu/nonlinear-dynamics-research/edm/) tools [cppEDM](https://github.com/SugiharaLab/cppEDM).

Functionality includes:
* Simplex projection (Sugihara and May 1990)
* Sequential Locally Weighted Global Linear Maps (S-map) (Sugihara 1994)
* Multivariate embeddings (Dixon et. al. 1999)
* Convergent cross mapping (Sugihara et. al. 2012)
* Multiview embedding (Ye and Sugihara 2016)

## Installation

You can install rEDM from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("rEDM")
```

OR from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SugiharaLab/rEDM")
```

## Example

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
par(mar = c(4, 4, 1, 1), mgp = c(2.5, 1, 0), oma = c(0, 0, 0, 0))
```

We begin by looking at annual time series of sunspots:

```{r sunspots}
df <- data.frame(yr = as.numeric(time(sunspot.year)), 
                 sunspot_count = as.numeric(sunspot.year))

plot(df$yr, df$sunspot_count, type = "l", 
     xlab = "year", ylab = "sunspots")
```

First, we use `EmbedDimension()` to determine the optimal embedding dimension, E:
```{r optimal-E}
library(rEDM)   # load the package

E.opt = EmbedDimension( dataFrame = df,    # input data
                        lib     = "1 280", # portion of data to train
                        pred    = "1 280", # portion of data to predict
                        columns = "sunspot_count",
                        target  = "sunspot_count" )
E.opt
```

Highest predictive skill is found between `E = 3` and `E = 6`. Since we generally want a simpler model, if possible, we use `E = 3` to forecast the last 1/3 of  data based on training (attractor reconstruction) from the first 2/3.

```{r simplex-projection}
simplex <- Simplex( dataFrame = df, 
                    lib     = "1   190", # portion of data to train
                    pred    = "191 287", # portion of data to predict
                    columns = "sunspot_count",
                    target  = "sunspot_count",
                    E       = 3 )

plot( df$yr, df$sunspot_count, type = "l", lwd = 2,
      xlab = "year", ylab = "sunspots")
lines( simplex$yr, simplex$Predictions, col = "red", lwd = 2)
legend( 'topleft', legend = c( "Observed", "Predicted (year + 1)" ),
        fill = c( 'black', 'red' ), bty = 'n', cex = 1.3 )
```

## Further Examples

Please see the package vignettes for more details:
```{r, eval = FALSE}
browseVignettes("rEDM")
```

### References
Sugihara G. and May R. 1990.  Nonlinear forecasting as a way of distinguishing 
chaos from measurement error in time series. Nature, 344:734–741.

Sugihara G. 1994. Nonlinear forecasting for the classification of natural 
time series. Philosophical Transactions: Physical Sciences and 
Engineering, 348 (1688) : 477–495.

Dixon, P. A., M. Milicich, and G. Sugihara, 1999. Episodic fluctuations in larval supply. Science 283:1528–1530.

Sugihara G., May R., Ye H., Hsieh C., Deyle E., Fogarty M., Munch S., 2012.
Detecting Causality in Complex Ecosystems. Science 338:496-500.
