---
title: 'Using rEDM to quantify time delays in causation'
author:
- Hao Ye
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Using rEDM to quantify time delays in causation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---
    
# Introduction
    
By default, convergent cross mapping (via the `ccm` function) tries to map between a lagged-coordinate vector from a "library" variable, $x$, and the simultaneous value of a target variable, $y$:
$$ y_{t+tp} = F\left(\vec{x}_t\right) = F\left(\langle x_t, x_{t-\tau}, \dots, x_{t-(E-1)\tau} \rangle \right) $$
where $tp = 0$.

However, note that `tp` is also an argument to `ccm`, and will accept both positive and negative values. This allows us to identify the ability to infer $F$ from the data at different values of `tp`, which is, to a first approximation, the time delay by which information about $y$ is encoded in the time series of $x$.

Note here that negative values of `tp` ($tp < 0$) indicate that *past* values of $y$ are best cross mapped from the reconstructed state of $\vec{x}$. This suggests a dynamical signal that appears first in $y$ and later in $x$, and is consistent with $y$ causing $x$.

If there is no causation in the reverse direction (i.e. $x$ does not cause $y$), then we would expect that CCM in the opposite direction:
$$ x_{t+tp} = G\left(\vec{y}_t\right) = G\left(\langle y_t, y_{t-\tau}, \dots, y_{t-(E-1)\tau} \rangle \right) $$
would be best at a positive values of `tp` ($tp > 0$).

*This presumes that the time series are sampled frequently enough relative to the causation that a time delay can be detected. If causation is synchronous or nearly so, then we may find the optimal value of `tp` to be $0$ in both directions.*

# Method

We can determine the optimal values of `tp` by testing different values within the `ccm` function. Conventionally, we do this with as much of the data as possible, to obtain the cleanest signal.

First, we grab some demo time series from the `block_3sp` data.frame:
```{r}
library(rEDM)
data(paramecium_didinium)
```

Setup the cross mapping runs we want to do:
```{r}
vars <- names(paramecium_didinium)[2:3] # c("paramecium", "didinium")

# generate all combinations of lib_column, target_column, tp
params <- expand.grid(lib_column = vars, 
                      target_column = vars, 
                      tp = -10:10)

# throw out cases where lib == target
params <- params[params$lib_column != params$target_column, ]

# E = 3 is optimal or very close to optimal for both vars
# In other circumstances, we should use the best univariate E for each lib_column
E <- 3
```

Perform cross mapping runs:
```{r, warning = FALSE}
output <- do.call(rbind, lapply(seq_len(NROW(params)), function(i) {
    ccm(paramecium_didinium, E = 3, 
        lib_sizes = NROW(paramecium_didinium), random_libs = FALSE, 
        lib_column = params$lib_column[i], 
        target_column = params$target_column[i], 
        tp = params$tp[i])
}))
```

# Results

Create an additional column to describe the direction of the cross mapping:
```{r}
output$direction <- paste(output$lib_column, "xmap to", output$target_column)
```

Plot the results (cross map skill vs. tp):
```{r, fig.width = 6}
library(ggplot2)
ggplot(output, aes(x = tp, y = rho, color = direction)) + 
    geom_line() + theme_bw()
```

### Causal Direction
As expected for this simple predator-prey system, we see evidence for causation in both directions (cross map skill, `rho`, peaks for negative `tp`).

### Strength of Causation
The strength of the interaction looks approximately equal (as inferred from cross map skill) -- though this inference has a weak assumption that all else is equal, including observational error; these assumptions are reasonable for this laboratory experiment.

### Time Delays
We further note that the estimated time delay for the effect of predators (didinium) on prey (paramecium) is fast ($tp$ is close to 0 for "paramecium xmap to didinium") compared to the effect of prey on predators ($tp$ is more negative for "didinium xmap to paramecium"). This is consistent with the ecological interpretation that changes in predator abundance affect prey abundance quickly, while there is a slower response of predator abundance to changes in prey abundance.
